---
title: "[mRNA] Transposition of the Great Arteries (TGA) with Systemic Right Ventricle (RV) and Systemic Left Ventricle (LV) DGE Analysis"
author: "Sara Tahir"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview
This pipeline analyzes the transcriptomic response of Transposition of the Great Arteries (TGA) in patients with systemic right and left ventricles, RV and LV, respectively. This dataset includes both mRNA and miRNA, and the analysis will include their interaction and sex stratification.

**Platform:** Agilent-072363 SurePrint G3 Human GE v3 8x60K Microarray 039494 [Feature Number Version]

**Design:** 16 Controls (11 males, 5 females), 16 TGA-LV patients (11 males, 5 females), 16 TGA-RV (11 males, 5 females)

## Setup and Read data
```{r package-setup, message = FALSE}
# Load/Install Manager
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, gplots, here, ggplot2,      # CRAN
  GEOquery, limma, edgeR, sva # Bioconductor
)
```

```{r read-data, message = FALSE}
# Read in geo dataset
geo_data <- getGEO(filename = "GSE215939_series_matrix.txt")
metadata <- pData(geo_data)

files <- list.files(path = "./GSE215939_RAW", pattern = "\\.txt$", full.names = TRUE)
x <- read.maimages(files, source="agilent", green.only = TRUE, other.columns="gIsWellAboveBG")
colnames(x$E) <- gsub("^.*/(GSM[0-9]+)_.*$", "\\1", colnames(x))
colnames(x$Eb) <- gsub("^.*/(GSM[0-9]+)_.*$", "\\1", colnames(x))
```
## Pre-Normalization
Save MA plots, view distribution before and after normalization
```{r pre-normalization}
plotMA3by2(x)

signal_data_frame <- stack(as.data.frame(log2(x$E)))
background_data_frame <- stack(as.data.frame(log2(x$Eb)))

ggplot(signal_data_frame) +
  geom_boxplot(aes(x = ind, y = values)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Signal")

ggplot(background_data_frame) +
  geom_boxplot(aes(x = ind, y = values))  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Background")
```
## Processing and Normalization
```{r normalization}
# Standard agilent microarray data processing
y <- backgroundCorrect(x, method="normexp")
y <- normalizeBetweenArrays(y, method="quantile")

Control <- y$genes$ControlType==1L
# The filterByExpr equivalent for Agilent
# Keep probes that are statistically above background in at least 5 patients (smallest group size)
smallestGroupSize <- 5 # Female group size is 5
IsExpr <- rowSums(y$other$gIsWellAboveBG > 0) >= smallestGroupSize
notProbe <- !grepl("^A_", y$genes$GeneName)
yfilt <- y[!Control & IsExpr &notProbe, ]

# Create boxplot for normalized data
signal_data_frame <- stack(as.data.frame(yfilt$E))
ggplot(signal_data_frame) +
  geom_boxplot(aes(x = ind, y = values)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Signal Normalized")

# View density plot before and after filtering
probe_means_clean <- rowMeans(yfilt$E)
probe_means_original <- rowMeans(y$E)
dens_clean <- density(probe_means_clean)
dens_original <- density(probe_means_original)

y_max <- max(dens_clean$y, dens_original$y)

plot(dens_original, 
     col = "black",
     lwd = 2,
     main = "Density of Genes Expressed Above Background", 
     xlab = "Log2 Intensity", 
     ylim = c(0, y_max))
lines(dens_clean, col = "red", lwd = 2)
legend("topright", 
       legend = c("Pre-Filter (All Probes)", "Post-Filter (Kept Probes)"),
       col = c("black", "red"), 
       lwd = 2, 
       lty = c(2, 1))

# Mean-Variance plot
pre_mean <- rowMeans(y$E)
pre_sd <- apply(y$E, 1, sd)

kept_mean <- rowMeans(yfilt$E)
kept_sd <- apply(yfilt$E, 1, sd)

plot(pre_mean, pre_sd, 
     pch = 16, cex = 0.5, col = rgb(0,0,0,0.1), # Transparent black dots
     main = "Mean-Variance Trend", 
     xlab = "Average Log2 Expression", 
     ylab = "Standard Deviation")

# Highlight the kept probes in blue
points(kept_mean, kept_sd, 
       pch = 16, cex = 0.5, col = rgb(0,0,1,0.1))

legend("topright", legend=c("Discarded Noise", "Kept Probes"), 
       col=c("black", "blue"), pch=16)

```
## Metadata formatting
```{r metadata-formatting}
accession_to_subject <- setNames( metadata$title, metadata$geo_accession)
colnames(yfilt) <- accession_to_subject[colnames(yfilt)]

# Turn TGA-LV to TGALV
metadata$source_name_ch1 <- gsub("-", "", metadata$source_name_ch1)
# Ensure data is aligned
metadata <- metadata[match(colnames(yfilt), metadata$title), ]
all(metadata$title == colnames(yfilt)) # Should return true
```

## Outlier Removal Toggle
```{r outlier-removal}
remove_outlier <- FALSE

if(remove_outlier) {
  target <- "TH_457"
  yfilt <- yfilt[, colnames(yfilt) != target]
  metadata <- metadata[metadata$title != target, ]
  
  message(paste("Outlier", target, "has been removed."))
} else {
  message("No samples removed.")
}
```

## Sample Quality Control
Used hierarchical clustering and PCA plots
``` {r sample-qc}
# Hierarchical clustering
dist_mat <- dist(t(yfilt$E))
hclust_res <- hclust(dist_mat)

group_labels <- paste(metadata$source_name_ch1, metadata$`gender:ch1`, metadata$title, sep = "_")

plot(hclust_res, 
     labels = group_labels,
     main = "Hierarchical Clustering of Samples", 
     xlab = "Samples", 
     sub = "", 
     cex = 0.6)

# Calculate variance for all genes across all samples
gene_vars <- apply(yfilt$E, 1, var, na.rm = TRUE)
# Grab top 500 genes
top_indices <- order(gene_vars, decreasing = TRUE)[1:500]
dist_mat_top <- dist(t(yfilt$E[top_indices, ]))
plot(hclust(dist_mat_top), 
     labels = group_labels, 
     main = "Hierarchical Clustering (Top 500 Variable Genes)", 
     cex = 0.6)

# Generate gene heatmap
heatmap_matrix <- yfilt$E[top_indices, ]
colnames(heatmap_matrix) <- group_labels
heatmap.2(heatmap_matrix, 
          trace = "none",         
          scale = "row",          
          col = colorRampPalette(c("blue", "white", "red"))(50),
          main = "Top 500 Variable Genes",
          cexCol = 0.6, 
          labRow = FALSE,         # CRITICAL VISUAL FIX: Hides the 500 overlapping row names
          margins = c(10, 5))

# PCA Plots
pca_res <- prcomp(t(yfilt$E), scale. = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_plot_data <- cbind(pca_df, metadata)

# Scree Plot
pca_res.var <- pca_res$sdev^2
pca_res.var.per <- round(pca_res.var/sum(pca_res.var)*100,1)
barplot(pca_res.var.per, main = "Scree Plot", xlab = "Principal Component", ylab = "Percent Variation")

# Create a combined grouping variable for the colors
pca_plot_data$Group <- paste(pca_plot_data$source_name_ch1, pca_plot_data$`gender:ch1`, sep = "_")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  
  # Assign complementary light/dark shades so the diseases group visually
  scale_color_manual(
    name = "Disease & Sex",
    values = c(
      "Control_Female" = "darkgray", 
      "Control_Male" = "black", 
      "TGALV_Female" = "deepskyblue", 
      "TGALV_Male" = "darkblue",
      "TGARV_Female" = "lightcoral",
      "TGARV_Male" = "firebrick"
    )
  ) +
  
  labs(
    title = "PCA Plot",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)")
  )
```

## Batch Correction
Use SVA
``` {r batch-correction}
group <- factor(paste(metadata$source_name_ch1, metadata$`gender:ch1`, sep = "_"))

mod  <- model.matrix(~ group, data = metadata)          # Full model (with intercept)
mod0 <- model.matrix(~ 1, data = metadata) # Null model (intercept only)

n_sv_conservative <- num.sv(yfilt$E, mod, method = "be")

message("The conservative BE method recommends: ", n_sv_conservative, " batches.")

svobj <- sva(yfilt$E, mod, mod0)

design <- model.matrix(~ 0 + group)
colnames(design) <- gsub("group", "", colnames(design))

if(svobj$n.sv > 0) {
  # If SVA found hidden batches, append them as new columns (SV1, SV2, etc.)
  design <- cbind(design, svobj$sv)
  colnames(design)[(ncol(design) - svobj$n.sv + 1):ncol(design)] <- paste0("SV", 1:svobj$n.sv)
  message(paste("SVA found and isolated", svobj$n.sv, "hidden batch effect(s)."))
} else {
  message("SVA found 0 hidden batches.")
}
```

## Post Batch Correction QC
```{r batch-correction-qc}
yfilt_clean <- removeBatchEffect(yfilt$E, 
                                 covariates = svobj$sv, 
                                 design = mod)
pca_clean <- prcomp(t(yfilt_clean), scale. = TRUE)
pca_df_clean <- as.data.frame(pca_clean$x)
pca_plot_data_clean <- cbind(pca_df_clean, metadata)

pca_plot_data_clean$Group <- paste(pca_plot_data_clean$source_name_ch1, 
                                   pca_plot_data_clean$`gender:ch1`, sep = "_")

# Plot it
ggplot(pca_plot_data_clean, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  scale_color_manual(
    name = "Disease & Sex",
    values = c("Control_Female" = "darkgray", "Control_Male" = "black", 
               "TGALV_Female" = "deepskyblue", "TGALV_Male" = "darkblue",
               "TGARV_Female" = "lightcoral", "TGARV_Male" = "firebrick")
  ) +
  labs(
    title = "PCA Plot (Post-SVA Batch Correction)",
    x = paste0("PC1 (", round(summary(pca_clean)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_clean)$importance[2,2] * 100, 1), "%)")
  )

# 3. Post-SVA Hierarchical Clustering
# Calculate the distance matrix on the clean data
dist_mat_clean <- dist(t(yfilt_clean))
hclust_clean <- hclust(dist_mat_clean)

# Plot the clean dendrogram
plot(hclust_clean, 
     labels = pca_plot_data_clean$Group,
     main = "Hierarchical Clustering (Post-SVA)", 
     xlab = "Samples", 
     sub = "", 
     cex = 0.6)

######## Just look at the top 500 genes
gene_vars_clean <- apply(yfilt_clean, 1, var, na.rm = TRUE)
top_indices_clean <- order(gene_vars_clean, decreasing = TRUE)[1:500]
yfilt_clean_top500 <- yfilt_clean[top_indices_clean, ]

pca_top500 <- prcomp(t(yfilt_clean_top500), scale. = TRUE)
pca_df_top500 <- as.data.frame(pca_top500$x)
pca_plot_data_top500 <- cbind(pca_df_top500, metadata)

# Ensure the Group column is formatted
pca_plot_data_top500$Group <- paste(pca_plot_data_top500$source_name_ch1, 
                                    pca_plot_data_top500$`gender:ch1`, sep = "_")

ggplot(pca_plot_data_top500, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  scale_color_manual(
    name = "Disease & Sex",
    values = c("Control_Female" = "darkgray", "Control_Male" = "black", 
               "TGALV_Female" = "deepskyblue", "TGALV_Male" = "darkblue",
               "TGARV_Female" = "lightcoral", "TGARV_Male" = "firebrick")
  ) +
  labs(
    title = "PCA Plot: Top 500 Most Variable Genes (Post-SVA)",
    x = paste0("PC1 (", round(summary(pca_top500)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_top500)$importance[2,2] * 100, 1), "%)")
  )

dist_mat_top500 <- dist(t(yfilt_clean_top500))
hclust_top500 <- hclust(dist_mat_top500)

plot(hclust_top500, 
     labels = pca_plot_data_top500$Group,
     main = "Hierarchical Clustering: Top 500 Genes (Post-SVA)", 
     xlab = "Samples", 
     sub = "", 
     cex = 0.6)

heatmap_matrix_clean <- yfilt_clean_top500
colnames(heatmap_matrix_clean) <- pca_plot_data_top500$Group

heatmap.2(heatmap_matrix_clean, 
          trace = "none",         
          scale = "row",          
          col = colorRampPalette(c("dodgerblue", "white", "firebrick"))(50),
          main = "Top 500 Variable Genes (Post-SVA)",
          cexCol = 0.6, 
          labRow = FALSE,  # Hides the 500 gene names so the plot isn't a black smudge       
          margins = c(10, 5))
```

## Limma DGE
```{r limma}
rownames(design) <- colnames(yfilt)

contrast_matrix <- makeContrasts(
  TGA_LV_Female_vs_Control_Female = TGALV_Female - Control_Female,
  TGA_LV_Male_vs_Control_Male = TGALV_Male - Control_Male,
  TGA_RV_Female_vs_Control_Female = TGARV_Female - Control_Female,
  TGA_RV_Male_vs_Control_Male = TGARV_Male - Control_Male,
  levels = design
)

# Fix heteroscedasticity
aw <- arrayWeights(yfilt, design)
barplot(aw, main="Array Weights per Patient", ylab="Weight", las=2, cex.names=0.5, names.arg = colnames(yfilt))
abline(h=1, lty=2, col="red")

fit <- lmFit(yfilt, design = design, weights = aw)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, robust = T)

# View Summaries
resultsSummary <- decideTests(fit2)
print(summary(resultsSummary))
vennDiagram(resultsSummary, cex = c(1, 0.6, 1))
```

## Post-Analysis and Result Generation
```{r post-analysis}
output_dir <- "mrna_results"
dir.create(output_dir, showWarnings = FALSE)

# Load the official HGNC list for validation (Keep this exactly as you had it)
hgnc_complete_set_df <- read_tsv("hgnc_complete_set.txt", show_col_types = FALSE) %>%
  dplyr::select(HGNC_Symbol = symbol)

contrasts_list <- colnames(contrast_matrix)

for (contrast_name in contrasts_list) {
  res_df <- topTable(fit2, coef = contrast_name, number = Inf) 
  
  # Clean and format the Agilent output
  res_df_clean <- res_df %>%
    dplyr::rename(HGNC_Symbol = GeneName) %>%               
    dplyr::filter(!is.na(HGNC_Symbol) & HGNC_Symbol != "")  
  
  # Validate against the official HGNC list
  joined_results <- res_df_clean %>%
    dplyr::inner_join(hgnc_complete_set_df, by = "HGNC_Symbol")
  
  # Handle duplicates: Sort by p-value, keep only the most biologically significant probe per gene
  final_results_table <- joined_results %>%
    dplyr::arrange(adj.P.Val) %>% 
    dplyr::distinct(HGNC_Symbol, .keep_all = TRUE)
  
  # Final formatting for the CSV
  final_results_table <- final_results_table %>%
    dplyr::select(HGNC_Symbol, logFC, P.Value, adj.P.Val) %>%
    dplyr::filter(!is.na(adj.P.Val))
  
  # Save to CSV
  file_name <- paste0(output_dir, "/dge_results_", contrast_name, "_mRNA_final.csv")
  write_csv(final_results_table, file_name)
  
  message(paste("Saved:", file_name))
}
```